Parameter Recovery Validation

### 2025-05-30 nbs/concepts/parameter-recovery-validation/index.qmd
docs: perform prior predictive before generating synthetic data for validation
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@

**Important**: For meaningful prior predictive checks, we must ensure the plotted data corresponds to the plotted parameters. This requires sampling parameters first, then generating data using those specific parameters.

# STEP 1: Sample parameters from the prior first
prior_parameter_samples = model.sample_system_parameters(
    num_samples=1000,
    constrain_to_pattern=False,  # Use full prior ranges
    n_genes=10,
    n_cells=200
)

# STEP 2: Generate data using a subset of those parameters for visualization
# Take first parameter sample for data generation
torch.manual_seed(1)
pyro.set_rng_seed(1)
single_param_sample = {key: value[:1] for key, value in prior_parameter_samples.items()}

prior_predictive_adata = model.generate_predictive_samples(
    num_cells=200,
    num_genes=10,
    samples=single_param_sample,  # Use specific parameters, not random sampling
    return_format="anndata"
)

# Add UMAP and clustering for visualization
sc.pp.neighbors(prior_predictive_adata, n_neighbors=10, random_state=42)
sc.tl.umap(prior_predictive_adata, random_state=42)
sc.tl.leiden(prior_predictive_adata, random_state=42)

# STEP 3: Plot with corresponding parameters and data
fig_prior = plot_prior_predictive_checks(
    model=model,
    prior_adata=prior_predictive_adata,        # Data generated from known parameters
    prior_parameters=prior_parameter_samples,  # Full parameter distribution
    figsize=(15, 10),
    save_path="reports/docs/prior_predictive",
)

>>>>

# If we wanted to generate data from a specific parameter set, we could
# first sample some parameters from the prior
# prior_parameter_samples = model.sample_system_parameters(
#     num_samples=1000,
#     constrain_to_pattern=False,  # Use full prior ranges
#     n_genes=10,
#     n_cells=200
# )
# and then use those to generate data from a single sample
# Take first parameter sample for data generation
# single_param_sample = {key: value[:1] for key, value in prior_parameter_samples.items()}
# And then set
# model.generate_predictive_samples(samples=single_param_sample, ...)
# But here we want to see the full breadth of what the prior can produce
# so we generate data from multiple parameter sets in a single call

prior_predictive_adata = model.generate_predictive_samples(
    num_cells=200,
    num_genes=10,
    num_samples=100,  # Generate data from 100 different parameter sets
    return_format="anndata"
)

# Sample parameters from the prior
prior_parameter_samples = model.sample_system_parameters(
    num_samples=1000,
    constrain_to_pattern=False,  # Use full prior ranges
    n_genes=10,
    n_cells=200
)

# Add UMAP and clustering though these will be uninformative in the context of prior predictive checks
sc.pp.neighbors(prior_predictive_adata, n_neighbors=10, random_state=42)
sc.tl.umap(prior_predictive_adata, random_state=42)
sc.tl.leiden(prior_predictive_adata, random_state=42)

# Plot randomly sampled parameters and data
fig_prior = plot_prior_predictive_checks(
    model=model,
    prior_adata=prior_predictive_adata,
    prior_parameters=prior_parameter_samples,
    figsize=(15, 10),
    save_path="reports/docs/prior_predictive",
)

@@
